---
title: "SQL Databases for Students and Educators"
author: "Pachá"
date: "2021-02-02"
output:
  rmdformats::downcute:    
    self_contained: false
    default_style: dark
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# About

<img src="https://opensource.nyc3.cdn.digitaloceanspaces.com/attribution/assets/PoweredByDO/DO_Powered_by_Badge_blue.svg" width="200">

Publicly accessible databases often impose query limits or require registration. Even when I maintain public and limit-free APIs, I never wanted to host a public database because I tend to think that the connection strings are a problem for the user.

I've decided to host different light/medium size by using PostgreSQL, MySQL and SQL Server backends (in strict descending order of preference!). 

Why 3 database backends? I think there are a ton of small edge cases when moving between DB back ends and so testing lots with live databases is quite valuable. With this resource you can benchmark speed, compression, and DDL types.

Please send me a tweet if you need the connection strings for your lectures or workshops. My Twitter username is [\@pachadotdev](https://twitter.com/pachadotdev).

# Intendo

It's a dummy database that originally we used at the [My Organization's First R Package](https://github.com/rstudio-conf-2020/my-org-first-pkg) and contains managerial information of a company called Intendo, which creates mobile games. The tables contain information for the 2015 year for all sorts of information for their only game: Super Jetroid. See the recent updated to this
database at [intendo](https://github.com/rich-iannone/intendo).

This database can be helpful for workshops and showing how to create an R package focused on the particular needs of an organization, including easier data access, shared functions for data transformation and analysis, and a common look and feel for reporting.

Check the structure of the tables in the next preview. The tables ending with
"_f" contain deliberated errors.

```{r, message=FALSE}
library(dplyr)
library(RPostgres)

con <- dbConnect(
  Postgres(),
  user = Sys.getenv("dbedu_usr"),
  password = Sys.getenv("dbedu_pwd"),
  dbname = "intendo",
  host = "databases.pacha.dev"
)

dbListTables(con)

tbl(con, "sj_user_summary_small") %>% glimpse()

dbDisconnect(con)
```

# Financial (aka Loan Application)

A database that contains Financial dataset contains 606 successful and 76 not successful loans along with their information and transactions.

This database comes from the [Relational Dataset Repository](https://relational.fit.cvut.cz/dataset/Financial) where you can find the database schema and more information. The database full name is "PKDD'99 Financial Dataset or Loan Application", and this particular version corresponds to Financial_ijs by Janez Kranjc.

Check the structure of the tables in the next preview.

```{r, message=FALSE}
library(dplyr)
library(RPostgres)

con <- dbConnect(
  Postgres(),
  user = Sys.getenv("dbedu_usr"),
  password = Sys.getenv("dbedu_pwd"),
  dbname = "financial",
  host = "databases.pacha.dev"
)

dbListTables(con)

tbl(con, "clients") %>% glimpse()
```

# SQL dumps

## Intendo

* [PostgreSQL](https://databases.pacha.dev/dumps/intendo-postgres.sql.gz) (15 MB, check the download with the [md5sum](https://databases.pacha.dev/dumps/intendo-postgres.sql.gz.md5))

* [MySQL/MariaDB](https://databases.pacha.dev/dumps/intendo-mysql.sql.gz) (16 MB, check the download with the [md5sum](https://databases.pacha.dev/dumps/intendo-mysql.sql.gz.md5))

* [SQL Server](https://databases.pacha.dev/dumps/intendo-sqlserver.sql.gz) (26 MB, check the download with the [md5sum](https://databases.pacha.dev/dumps/intendo-sqlserver.sql.gz.md5))

## Financial

* [PostgreSQL](https://databases.pacha.dev/dumps/financial-postgres.sql.gz) (11 MB, check the download with the [md5sum](https://databases.pacha.dev/dumps/financial-postgres.sql.gz.md5))

* [MySQL/MariaDB](https://databases.pacha.dev/dumps/financial-mysql.sql.gz) (178 MB, check the download with the [md5sum](https://databases.pacha.dev/dumps/financial-mysql.sql.gz.md5))

* SQL Server: Coming Soon

<!-- * [SQL Server](https://databases.pacha.dev/dumps/financial-sqlserver.sql.gz) (269 MB, check the download with the [md5sum](https://databases.pacha.dev/dumps/financial-sqlserver.sql.gz.md5)) -->

# Using the dumps locally

## Restore dumps

### PostgreSQL

<!-- pg_dump intendo > intendo-postgres.sql -->
 
```bash
sudo -i -u postgres
gunzip -c intendo-postgres.sql.gz | psql intendo
```

### MySQL/MariaDB

<!-- mysqldump intendo > intendo-mysql.sql -->

From SQL:

``` SQL
# mysql 
CREATE DATABASE intendo;
```

Then from bash:
```bash
sudo su
gunzip intendo-mysql.sql.gz
mysql intendo < intendo-mysql.sql
```

### SQL Server

<!-- BACKUP DATABASE intendo TO DISK = 'filepath';  -->

From bash:
```bash
gunzip intendo-mysql.sql.gz
sqlcmd -S localhost -U SA
```

Then in SQL:
```SQL
RESTORE DATABASE intendo FROM DISK = 'intendo-sqlserver.sql'
GO
```

## Create users

You can create a generic user (let's say `student`) and grant read-only access.

### PostgreSQL

```SQL
# sudo -i -u postgres 
# psql -d intendo
CREATE ROLE student;
CREATE ROLE teacher;
ALTER ROLE student WITH PASSWORD 'SomePassword';
ALTER ROLE teacher WITH PASSWORD 'SomePassword';
ALTER ROLE student WITH LOGIN;
ALTER ROLE teacher WITH LOGIN;

GRANT CONNECT ON DATABASE intendo TO student;
GRANT USAGE ON SCHEMA public TO student;
GRANT SELECT ON ALL TABLES IN SCHEMA public TO student;
GRANT SELECT ON ALL SEQUENCES IN SCHEMA public TO student;
ALTER DEFAULT PRIVILEGES IN SCHEMA public GRANT SELECT ON TABLES TO student;
REVOKE CREATE ON SCHEMA public FROM public;

GRANT CONNECT ON DATABASE intendo TO teacher;
GRANT USAGE ON SCHEMA public TO teacher;
GRANT SELECT ON ALL TABLES IN SCHEMA public TO teacher;
GRANT SELECT ON ALL SEQUENCES IN SCHEMA public TO teacher;
ALTER DEFAULT PRIVILEGES IN SCHEMA public GRANT SELECT ON TABLES TO teacher;
GRANT CREATE ON SCHEMA public to teacher;
```

### MySQL/MariaDB

```SQL
# sudo mysql
USE intendo;

CREATE USER student IDENTIFIED BY 'SomePassword';
CREATE USER teacher IDENTIFIED BY 'SomePassword';

GRANT SELECT ON intendo.* TO student;

GRANT SELECT ON intendo.* TO teacher;
GRANT CREATE ON intendo.* TO teacher;
GRANT INSERT ON intendo.* TO teacher;
GRANT ALTER ON intendo.* TO teacher;
```

### SQL Server

```SQL
# sqlcmd -S localhost -U SA

USE intendo
GO

CREATE LOGIN student WITH PASSWORD = 'SomePassword'
GO
CREATE USER student FOR LOGIN student
GO
GRANT SELECT ON SCHEMA :: dbo TO student
GO

CREATE LOGIN teacher WITH PASSWORD = 'SomePassword'
GO
CREATE USER teacher FOR LOGIN teacher
GO
GRANT SELECT ON SCHEMA :: dbo TO teacher
GO
GRANT CREATE TABLE TO teacher
GO
GRANT INSERT ON SCHEMA :: dbo TO teacher
GO
GRANT ALTER ON SCHEMA :: dbo TO teacher
GO
GRANT UPDATE ON SCHEMA :: dbo TO teacher
GO

# You'll also want to limit SQL Server memory
# This avoids strange crashes
sp_configure 'show advanced options', 1;
GO
RECONFIGURE;
GO
sp_configure 'max server memory', 1024;
GO
RECONFIGURE;
GO
```

## Allow external connections

For this project I had to allow connections from outside the server.

### Postgres (version 12)

1. Open `/etc/postgresql/12/main/postgresql.conf`, comment the line that says `listen_addresses = 'localhost'`,
add `listen_address = '*'` below it.
2. Open `/etc/postgresql/13/main/pg_hba.conf`, paste these lines at the end
```bash
# TYPE  DATABASE        USER    ADDRESS         METHOD
host    all             all     0.0.0.0/0       md5
host    all             all     :/0             md5
```
3. Run `sudo systemctl restart postgres`.
4. Open the port with `sudo ufw allow 5432/tcp`.

### MariaDB (i.e. MySQL replacement)

1. Open `/etc/mysql/mariadb.conf.d/50-server.cnf`, comment the line that says `bind-address = 127.0.0.1`.
2. Run `sudo systemctl mysql`.
3. Open the port with `sudo ufw allow 3306/tcp`.

### SQL Server

3. Access as administrator and run
```SQL
EXEC sys.sp_configure N'remote access', N'1'
GO
RECONFIGURE WITH OVERRIDE
GO
```
2. Open the ports with 
```bash
sudo ufw allow 1433/tcp
sudo ufw allow 4022/tcp
sudo ufw allow 135/tcp
sudo ufw allow 1434/tcp
sudo ufw allow 1434/udp
```

# Problems with MySQL/Maria SQL Server

## MySQL/MariaDB

From R, the RMariaDB package copies around 20 times slower than RMySQL. Also using odbc for this SQL backend is not very efficient. Just use RMySQL even if (by any reason) you prefer it over MySQL.

When you backup, at least with MariaDB, `mysqldump intendo > intendo-mysql.sql` is hundreds of times faster than `mysql intendo > intendo-mysql.sql`.

## SQL Server

Beware that SQL Server shows strange problems with Linux and Mac! For example, when I connect from a Mac, I experience random disconnections and the next error message:
```bash
nanodbc/nanodbc.cpp:4483: 00000: [Microsoft][ODBC Driver 18 for SQL Server]Communication link failure 
```
With Ubuntu I get this another error message and RStudio crashes:
```bash
terminate called after throwing an instance of 'nanodbc::database_error'
```
Therefore, my preference for DB backends is not arbitrary.

I had to divide my data and write in chunks of 50,000 rows or it just disconnects. PostgreSQL and MariaDB can write tables containing two million rows right away.

# Cite this work

This work is licensed under [Creative Commons  Attribution 4.0 International (CC BY 4.0) ](https://creativecommons.org/licenses/by/4.0/).

BibTeX entry:
```
@misc{databases_pacha,
  title = {SQL Databases for Students and Educators},
  url = {https://databases.pacha.dev/},
  author = {Vargas, Mauricio},
  doi = {10.5281/zenodo.4136985},
  publisher = {Self-published},
  year = {2021},
  month = {Feb},
  note = {Accessed: Month DD, YYYY}
}
```

# Access and support

If you need access to the online DBs (i.e., not configuring your local copy), 
send me an email to *ma vargas 11 at uc dot cl* or a tweet to 
[\@pachadotdev](https://twitter.com/pachadotdev)).

I also have a [Buy me a coffee](https://www.buymeacoffee.com/pacha) profile. If you like my work, buy me a coffee, it shall be used to produce more of it. Thank you for your support!
