---
title: "SQL Databases for Students and Educators"
author: "Pachá"
date: "2024-04-08"
output:
  rmdformats::downcute:    
    self_contained: false
    default_style: dark
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

*Update 2022-12-12: I'm using just PostgreSQL. I just had too many heaches
with MariaDB/MySQL. Also, the server is quite modest and I prefer to provide 
good PostgreSQL access than bad access to PostgreSQL and MariaDB.*

*Update 2022-09-30: The SQL Server educational license expired. From now on,
only PostgreSQL and MariaDB databases will be available.*

# About

Publicly accessible databases often impose query limits or require registration. 
Even when I maintain public and limit-free APIs, I never wanted to host a public 
database because I tend to think that the connection strings are a problem for 
the user.

I've decided to host different light/medium size by using PostgreSQL, MariaDB
and SQL Server. I tend to  prefer PostgreSQL because it easier to configure on
client and server side, without strange tricks (see the *Problems with
MySQL/MariaDB and SQL Server* section).

There are a ton of small edge cases when moving between DB back ends and so
testing lots with live databases is quite valuable. With this resource you can
benchmark speed, compression, and DDL types.

If you need access to the databases or need a SQL dump , send me an email to 
m.sepulveda+spam@mail.utoronto.ca (replace with +databases for a faster reply).

You can contribute to keep it up with the hosting costs here:

<script type="text/javascript" src="https://cdnjs.buymeacoffee.com/1.0.0/button.prod.min.js" data-name="bmc-button" data-slug="pacha" data-color="#FFDD00" data-emoji=""  data-font="Cookie" data-text="Buy me a coffee" data-outline-color="#000000" data-font-color="#000000" data-coffee-color="#ffffff" ></script>

# Available Databases

## Intendo

It's a dummy database that originally we used at the [My Organization's First R Package](https://github.com/rstudio-conf-2020/my-org-first-pkg) and contains 
managerial information of a company called Intendo, which creates mobile games. 
The tables contain information for the 2015 year for all sorts of information 
for their only game: Super Jetroid. See the recent updated to this
database at [intendo](https://github.com/rich-iannone/intendo).

This database can be helpful for workshops and showing how to create an R 
package focused on the particular needs of an organization, including easier 
data access, shared functions for data transformation and analysis, and a common
look and feel for reporting.

Check the structure of the tables in the next preview. The tables ending with
"_f" contain deliberated errors.

```{r, message=FALSE}
library(dplyr)
library(RPostgres)

con <- dbConnect(
  Postgres(),
  user = Sys.getenv("dbedu_usr"),
  password = Sys.getenv("dbedu_pwd"),
  dbname = "intendo",
  host = "databases.pacha.dev"
)

dbListTables(con)

tbl(con, "sj_user_summary_small") %>%
  glimpse()

dbDisconnect(con)
```

# Problems with MySQL/Maria and SQL Server

## MySQL/MariaDB

From R, the RMariaDB package copies around 20 times slower than RMySQL. Using 
odbc for this SQL backend is not very efficient. Just use RMySQL.

When you backup, at least with MariaDB, `mysqldump intendo > intendo-mysql.sql` 
is hundreds of times faster than `mysql intendo > intendo-mysql.sql`.

## SQL Server

Beware that SQL Server shows strange problems with Linux and Mac!

When I connect from a Mac, I experience random disconnections and the next error
message:

```bash
nanodbc/nanodbc.cpp:4483: 00000: [Microsoft][ODBC Driver 18 for SQL Server]Communication link failure 
```

With Ubuntu I get this another error message and RStudio crashes:

```bash
terminate called after throwing an instance of 'nanodbc::database_error'
```

I had to divide my data and write in chunks of 50,000 rows or it just 
disconnects. PostgreSQL and MariaDB can write tables containing two million rows
right away.

Therefore, my preference for PostgreSQL is not arbitrary.

# Cite this work

This work is licensed under [Creative Commons  Attribution 4.0 International (CC BY 4.0) ](https://creativecommons.org/licenses/by/4.0/).

BibTeX entry:
```
@misc{databases_pacha,
  title = {SQL Databases for Students and Educators},
  url = {https://databases.pacha.dev/},
  author = {Vargas, Mauricio},
  doi = {10.5281/zenodo.4136985},
  publisher = {Self-published},
  year = {2021},
  month = {Feb},
  note = {Accessed: Month DD, YYYY}
}
```
